---
title: "Correlation workflow"
author: "Noor Sohail"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
# Turn off Warnings and other console output messages from the whole document
```

```{r, cache=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(glue)
library(Seurat)
library(pheatmap)
library(devtools)
library(gridExtra)

library(CSCORE)
library(WGCNA)
library(SAVER)

ggplot2::theme_set(theme_light(base_size = 11))
opts_chunk[["set"]](
    cache = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE)
```

# User define values

```{r}
path_seurat <- "../data/seurat.RDS"
path_outs <- "05.0_workflow"

col_sample <- "sample"
col_celltype <- "celltype"
ct <- "B cells"

# Filtration parameters
filter <- TRUE
min_exp <- 0.2
min_cells <- 40
# min_perc <- 0.2
min_perc <- 0.5
```

- Path seurat: `r path_seurat`
- Metadata column with celltypes: `r col_celltype`
- Celltype to subset to: `r ct`
- Filter genes based on expression and frequency: `r filter`

List of all genes:
```{r}
# Fill in the list of genes you are interested in calculating correlations for
corr_genes_all <- c()

corr_genes_all <- read.csv("../data/surfaceome_genes.csv")$x
as.data.frame(corr_genes_all)
```

Starting off with `r length(corr_genes_all)` genes of interest.

# Load seurat object

```{r}
# Manual addition
load(bzfile("../data/seurat_integrated.RData.bz2"))

seurat <- seurat_integrated
DefaultAssay(seurat) <- "RNA"
seurat[["SCT"]] <- NULL
seurat[["integrated"]] <- NULL

rm(seurat_integrated)

Idents(object = seurat) <- "integrated_snn_res.0.8"
seurat <- RenameIdents(object = seurat, 
                               "0" = "Naive or memory CD4+ T cells",
                               "1" = "CD14+ monocytes",
                               "2" = "Activated T cells",
                               "3" = "CD14+ monocytes",
                               "4" = "Stressed cells / Unknown",
                               "5" = "CD8+ T cells",
                               "6" = "Naive or memory CD4+ T cells",
                               "7" = "B cells",
                               "8" = "NK cells",
                               "9" = "CD8+ T cells",
                               "10" = "FCGR3A+ monocytes",
                               "11" = "B cells",
                               "12" = "NK cells",
                               "13" = "B cells",
                               "14" = "Conventional dendritic cells",
                               "15" = "Megakaryocytes",
			       "16" = "Plasmacytoid dendritic cells")
seurat$celltype <- Idents(seurat)

DefaultAssay(seurat) <- "RNA"
```

```{r}
# library(SeuratDisk)
# seurat <- readRDS(path_seurat)

# Saving celltype information as column named celltype
# In order to use subset function later
seurat$celltype <- seurat@meta.data[col_celltype]
```


```{r}
Idents(seurat) <- col_celltype
DimPlot(seurat) + ggtitle("Celltypes")
```

```{r}
seurat@meta.data %>%
        ggplot() +
        geom_bar(aes(
            x = get(col_celltype),
            fill = get(col_celltype)),
            stat = "count", color = "black") +
        theme_classic() +
        NoLegend() +
        xlab("Celltype") +
        ylab("Number of Cells") +
        ggtitle("Celltypes") +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_text(aes(x = get(col_celltype), label = after_stat(count)), stat = "count", vjust = -0.5) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Subset to `r ct` cells

```{r}
seurat <- subset(seurat, celltype == ct)

# Removing genes that have 0 counts across all cells of the celltype
raw_rna <- GetAssayData(object =  seurat[['RNA']], layer = 'counts')
genes.use <- rowSums(raw_rna) > 0
genes.use <- names(genes.use[genes.use])
seurat <- seurat[genes.use, ]

n_cells <- ncol(seurat)
```

Working with `r n_cells` cells.

```{r}
seurat@meta.data %>%
        ggplot() +
        geom_bar(aes(
            x = get(col_sample),
            fill = get(col_sample)),
            stat = "count", color = "black") +
        theme_classic() +
        NoLegend() +
        ggtitle(glue("{ct} cells: Sample distribution")) +
        ylab("Number of Cells") +
        xlab("Sample") +
        theme(plot.title = element_text(hjust = 0.5)) +
        geom_text(aes(x = get(col_sample), label = after_stat(count)), stat = "count", vjust = -0.5) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Subset genes of interest

From the genes that were supplied, removing any that are not expressed in this dataset.

```{r}
corr_genes <- corr_genes_all[corr_genes_all %in% genes.use]
as.data.frame(corr_genes)
```

Next we look at the basic distribution of remaining genes of interest in terms of expression and number of cells they are expressed in.

These are the following filtration metrics that are set:

- Average expression < `r min_exp`
- Number of cells expressed in > `r min_cells`
- Percentage of cells expressed in > `r min_perc`

Filtration parameter was set to `r filter`. If FALSE, no further filtration will be done at this step.

```{r}
data_rna <- FetchData(seurat[["RNA"]], vars = corr_genes)

# Number of cells a gene is expressed in
num_cells <- colSums(data_rna > 0)
# Percentage of cells a gene is expressed in
perc_cells <- num_cells / ncol(seurat)
# Average expression of a gene
avg_expression <- colMeans(data_rna)

df_genes <- data.frame(num_cells, perc_cells, avg_expression)
df_genes <- df_genes %>% mutate(filter = !((perc_cells > min_perc) & (num_cells > min_cells) & (avg_expression > min_exp)))
df_genes$gene <- row.names(df_genes)
df_genes <- df_genes %>% arrange(desc(avg_expression), desc(perc_cells))

df_genes
```

```{r}
df_genes %>% ggplot() +
    geom_point(aes(x = perc_cells, y = avg_expression, color = filter)) +
    theme_classic()
```

```{r}
if (filter == TRUE) {
    corr_genes <- (df_genes %>% subset(filter == FALSE))$gene
}
```

`r length(corr_genes)` genes of interest remaining.

# Imputation and normalization

We compare three alternative methods of estimating expression levels to log normalization and assess their ability to account for dropout.

1. SCTransform (raw counts -> normalized counts)
2. MAGIC (raw counts -> imputed, normalized counts)
3. SAVER (raw counts -> imputed, normalized counts)

```{r}
# Get raw counts
raw_rna <- LayerData(seurat, assay = "RNA", layer = "counts")

# LOG NORMALIZATION
data_rna <- LayerData(seurat, assay = "RNA", layer = "data")

# SCT
# Re-run SCT on subset data
seurat <- SCTransform(seurat, return.only.var.genes = FALSE, min_cells = 1)
data_sct <-  LayerData(seurat, assay = "SCT", layer = "data")

# MAGIC
# Load conda environment
# myenvs <- reticulate::conda_list()
envname <- "merck"
reticulate::use_condaenv(envname, required = TRUE)
library(Rmagic)
data_magic <- magic(t(raw_rna), genes = corr_genes)$result
data_magic <- t(data_magic)

# SAVER
# Generate SAVER predictions for those genes
genes.ind <- which(rownames(raw_rna) %in% corr_genes)
data_saver <- saver(raw_rna, pred.genes = genes.ind, pred.genes.only = TRUE, estimates.only = TRUE, ncores = 8)
```

```{r}
# Store output so we don't have to re-run previous step each time
filename <- glue("{path_outs}/data_list.RDS")

# Put all matrices in a list for easy access
data_list <- list(data_rna, data_sct, data_magic, data_saver)
names(data_list) <- c("rna", "sct", "magic", "saver")
# saveRDS(data_list, filename)

# data_list <- readRDS(filename)
```

## Average expression for each method

```{r}
df_avg <- data.frame(gene = corr_genes)
for (method in names(data_list)) {
    data <- data_list[[method]]
    avg <- data.frame(rowMeans(data))
    colnames(avg) <- method
    avg$gene <- row.names(avg)

   df_avg <-  left_join(df_avg, avg, by = "gene")
}

pheatmap(df_avg %>% column_to_rownames(var = "gene"), scale = "column",
            cluster_col = TRUE, cluster_row = TRUE, show_rownames = FALSE)
```

# Correlation Estimates

We have a few different ways to compute correlation scores with their associated p-values:

1. Spearman correlation 
  - SCTransform counts -> spearman correlation matrix
  - MAGIC imputed -> spearman correlation matrix
  - SAVER imputed -> spearman correlation matrix
2. CS-CORE 
    - Raw RNA counts -> co-expression matrix

```{r}
# Compute spearman correlation for each method (except CS-CORE which is run later)
# Unique combination of each gene pair
genes_comb <- data.frame(t(combn(corr_genes, 2)))
n_comb <- nrow(genes_comb)
methods <- c("rna", "sct", "magic", "saver")

# Create dataframe with correlation and p-values scores
df_corr <- genes_comb %>% rename("Var1" = X1, "Var2" = X2)
df_corr[methods] <- NA
df_p_val <- genes_comb %>% rename("Var1" = X1, "Var2" = X2)
df_p_val[methods] <- NA

for (idx in 1:n_comb) {

    if (idx %% 200 == 0) {
        print(glue("{idx}/{n_comb} correlations computed."))
    }

    # Name of genes to run correlation on
    gene_1 <- genes_comb[idx, 1]
    gene_2 <- genes_comb[idx, 2]

    for (method in methods) {
        gene_1_exp <- data_list[[method]][gene_1, ]
        gene_2_exp <- data_list[[method]][gene_2, ]

        if (all(gene_1_exp == 0) | all(gene_2_exp == 0)) {
            # If a gene has no expression, set correlation = 0 and p-value = 1
            corr_val <- 0.0
            p_val <- 1.0
        } else {
            # Calculate spearman correlation and p-value otherwise
            tmp <- cor.test(gene_1_exp, gene_2_exp, method = "spearman", exact = FALSE)
            corr_val <- as.numeric(unname(tmp$estimate))
            p_val <- as.numeric(tmp$p.value)
        }

        # Store correlation and p-values
        df_corr[idx, method] <- corr_val
        df_p_val[idx, method] <- p_val
    }
}

# Run CS-CORE
CSCORE_result <- CSCORE(seurat, genes = corr_genes)

# Store CS-CORE results
tmp <- reshape2::melt(as.matrix(CSCORE_result$est)) %>% rename(cscore = value)
df_corr <- left_join(df_corr, tmp)
tmp <- reshape2::melt(as.matrix(CSCORE_result$p_value)) %>% rename(cscore = value)
df_p_val <- left_join(df_p_val, tmp)
```

## Heatmap of correlation estimates

Showing the patterns of correlation for each method. The x-axis and y-axis are the genes of interest with the corresponding correlation value for the pair as the value. Keep in mind that this is symmetric matrix.

```{r results = "asis"}
library(RColorBrewer)
methods <- c("rna", "sct", "magic", "saver", "cscore")

for (method in methods) {
    corr <- df_corr[c("Var1", "Var2", method)]
    corr_cp <- corr %>% rename(Var1 = Var2, Var2 = Var1)
    corr <- rbind(corr, corr_cp)
    mtx <- reshape2::dcast(corr, Var2 ~ Var1) %>% column_to_rownames("Var2")

    # Set the diagonal values: Correlation = 1, p-value = 1
    mtx <- as.matrix(mtx)
    diag(mtx) <- 1

    breaks <-  seq(-1, 1, by = 0.1)
    color <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaks))


    p <- pheatmap(mtx, show_rownames = FALSE, show_colnames = FALSE,
                color = color, breaks = breaks, silent = FALSE, main = method)
    knit_print(p)
}
```

# Compare correlation estimates across methods

Comparing the correlation scores for each gene pair for MAGIC, SAVER, and CS-CORE.

In these scatterplots, the gene-pairs that are colored red have different results for significance.

```{r fig.width=15}
methods <- c("magic", "saver", "cscore")
methods_comb <- data.frame(t(combn(methods, 2)))
plot_list <- list()

for (idx in 1:nrow(methods_comb)) {
    method_1 <- methods_comb[idx, 2]
    method_2 <- methods_comb[idx, 1]

    corr <- df_corr[c("Var1", "Var2", method_1, method_2)]
    p_val <- df_p_val[c("Var1", "Var2", method_1, method_2)]
    corr$sig_1 <- p_val[[method_1]]
    corr$sig_2 <- p_val[[method_2]]

    corr <- corr %>% mutate(sig = (sig_1 < 0.5) & (sig_2 < 0.05))

    p <- ggplot(corr) +
            geom_point(aes(x = get(method_1), y = get(method_2), color = sig)) +
            theme_classic() +
            NoLegend() +
            scale_color_manual(values = c("FALSE" = "red", "TRUE" = "black")) +
            labs(x = method_1, y = method_2, title = paste(method_1, "vs", method_2)) +
            theme(plot.title = element_text(size=rel(2))) +
            ylim(-1, 1) + xlim(-1, 1) +
            geom_abline(slope = 1, intercept = 0, color = "blue")

    plot_list[[idx]] <- ggplotGrob(p)

}

grid.arrange(grobs = plot_list, ncol = 3)
```

# SAVER and CS-CORE results

Based upon previous results, the results from CS-CORE and SAVER tend to agree with one another which is what we will use to generate the final list of gene pairs.

To make the most informed decision about which gene pairs to use, the final dataframe includes the following information:

- Gene1 and Gene2 (`Var1` and `Var2`)
- SAVER spearman correlation score (`saver`) and p-value (`p_saver`)
- CS-CORE correlation estimate (`cscore`) and p-value (`p_cscore`)
- Average expression for each gene in the pair after SAVER imputation (`Var1_avg` and `Var2_avg`)
- Ratio of average expression (gene 1 average / gene 2 average) (`avg_ratio`)
- Percent cells that express both genes before imputation (`perc_both`) to assess dropout

```{r}
df_results <- df_corr %>% select(Var1, Var2, saver, cscore)

# Including symmetric values
# This will make it easier to see which genes have large differences in average expression
corr_cp <- df_results %>% rename(Var1 = Var2, Var2 = Var1)
df_results <- rbind(df_results, corr_cp)

# Adding p-values for CS-CORE and SAVER
p_corr <- df_p_val %>% select(Var1, Var2, saver, cscore)
colnames(p_corr) <- c("Var1", "Var2", "p_saver", "p_cscore")
p_corr_cp <- p_corr %>% rename(Var1 = Var2, Var2 = Var1)
p_corr <- rbind(p_corr, p_corr_cp)
df_results <- merge(df_results, p_corr)

# Add average SAVER imputed values for each gene
idx <- match(df_results$Var1, df_avg$gene)
df_results$Var1_avg <- df_avg[idx, "saver"]
idx <- match(df_results$Var2, df_avg$gene)
df_results$Var2_avg <- df_avg[idx, "saver"]

# Ratio of average expression values
df_results <- df_results %>% mutate(avg_ratio = Var1_avg / Var2_avg)

data <- as.data.frame(t(data_list[["rna"]]))
df_results$perc_both <- NA
# Percent of cells that express both genes
for (idx in 1:nrow(df_results)) {
    gene_1 <- df_results[idx, "Var1"]
    gene_2 <- df_results[idx, "Var2"]

    data_sub <- data[c(gene_1, gene_2)]
    colnames(data_sub) <- c("Var1", "Var2")
    data_sub <- data_sub %>% subset(!(Var1 == 0) & !(Var2 == 0))
    df_results[idx, "perc_both"] <- nrow(data_sub) / n_cells
}

df_results
```

